Iterative Reduction: Parsing and Interpreting Code Without Trees

### 2025-01-18 Paper/paper.tex
Fixed proof on REG<BR, added BR(n)

>>>>

\section{$\beta$-Reducible Languages}

In this section we will try to understand better the class of languages which can be parsed using $\beta$-reduction.
The definition of $\beta$-reduction in this section will differ slightly from the remainder of this paper, but core ideas will be preserved.
We will construct a hierarchy of $\beta$-reducible languages, which we will compare with the Chomsky Hierarchy.

\subsection{$\beta$-Reducible Languages}

The simplest class in our hierarchy will be $\beta$-reducible languages, the class of which we will denote by \BR.
Define a \BR-schema to be a tuple $B=(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\hb)$ where
\begin{enumerate}
    \item $\Sigma$ is a finite alphabet of terminals,
    \item $\Gamma$ is a finite alphabet of non-terminals (generally $\Sigma$ and $\Gamma$ are disjoint, but we don't require this),
    \item $\square,\blacksquare$ are symbols not in $\Sigma,\Gamma$,
    \item $\bullet$ is a symbol (which may be in $\Sigma,\Gamma$) or the empty word,
    \item $s$ is a state, which is a map $\Sigma\to\Gamma$,
    \item $\hb$ is an initial $\beta$-reducer, which is a partial function $\hb\colon\Gamma\times(\Gamma\cup\set{\epsilon,\blacksquare})\partial\Gamma\cup\set{\epsilon,\bullet}$.
\end{enumerate}
Given a \BR-schema, we define a pre-$\beta$-reducer to be a function $\beta'\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given a string $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta'(\xi)$ as follows:
\begin{enumerate}
    \item If $\xi=\bullet$, then $\beta'(\xi)=\xi$
    \item If $\xi=\sigma\in\Gamma$, then $\beta'(\xi)=\hb(\sigma,\blacksquare)$ if it is defined, and $\sigma$ otherwise,
    \item If $\xi=\sigma_1\sigma_2\xi'$, then
        \begin{enumerate}
            \item $\beta'(\xi)=\hb(\sigma_1,\sigma_2)\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\hb(\sigma_1,\epsilon)\sigma_2\xi'$ if it is defined, otherwise
            \item $\beta'(\xi)=\sigma_1\beta'(\sigma_2\xi')$.
        \end{enumerate}
\end{enumerate}
And we define a $\beta$-reducer to be a function $\beta\colon(\Gamma\cup\set\bullet)^*\longto(\Gamma\cup\set\bullet)^*$ as follows: given $\xi\in(\Gamma\cup\set\bullet)^*$ we define $\beta(\xi)$ as follows:
\begin{enumerate}
    \item if $\xi=\epsilon$ and $\hb(\square,\blacksquare)=\tau$ is defined then $\beta(\xi)=\tau$,
    \item if $\xi=\sigma\xi'$ and $\hb(\square,\sigma)=\tau$ is defined then $\beta(\xi)=\tau\xi'$,
    \item otherwise $\beta(\xi)=\beta'(\xi)$.
\end{enumerate}
Then we define $\beta^*(\xi)$ to be the fixed point of successive applications of $\beta$ to $\xi$, i.e. $\beta^*(\xi)=\beta^n(\xi)$ where $\beta^n(\xi)=\beta^{n+1}(\xi)$.
For a word $w=\sigma_1\cdots\sigma_n=\in\Sigma^*$, we define $s(w)=s(\sigma_1)\cdots s(\sigma_n)\in\Gamma^*$.
The language \textit{generated} (or \textit{accepted}) by a \BR-schema $B$ is defined to be all the words $w\in\Sigma^*$ such that $\beta^*(s(w))=\bullet$.

Finally, let $\sigma\in\Gamma$: define the set $C(\sigma)=\bigcup_{n\geq1}C_n(\sigma)$ where $C_0(\sigma)=\set\sigma$ and $C_{n+1}(\sigma)=\set{\hb(\tau,\epsilon),\hb(\tau,\square)}[\tau\in C_n(\sigma)]$.
Call a \BR-schema \textit{valid} if $\sigma\notin C(\sigma)$ for all $\sigma\in\Gamma$ and $\hb(\square,\blacksquare)$ is either undefined or $\bullet$.
A $\beta$-reducible language is one generated by a valid \BR-schema; the class of all $\beta$-reducible languages is denoted \BR.
The reason for requiring a \BR-schema to be valid is to ensure that the $\beta$-reducer halts.
Otherwise, suppose $\sigma\in C(\sigma)$, so we could have that $\hb(\sigma,\square)=\sigma$ for example, and then $\beta$ would never halt on the string $\sigma$.
Validity is not a strong requirement.

According to theorem \ref{thrm:complexity}, all $\beta$-reducible languages can be reduced by their $\beta$-reducer in $O(n)$ time.

For example, $\set{a^nb^n}[n\geq1]$ is in \BR.
We define $B=(\Sigma,\Gamma,s,\hb)$ as follows:
\begin{enumerate}
    \item $\Sigma=\set{a,b}$,
    \item $\Gamma=\set{S_a,S_b,S_{ab},S_{aab}}$,
    \item $s(a)=S_a$ and $s(b)=S_b$,
    \item $\hb$ is defined by the rules:
        \[ S_a,S_b\to S_{ab},\quad S_a,S_{ab}\to S_{aab},\quad S_{aab},S_b=S_{ab},\quad S_{ab},\blacksquare\to\bullet \]
\end{enumerate}
Indeed, a word of the form $a\cdots ab\cdots b$ will be mapped to $S_a\cdots S_aS_b\cdots S_b$, which will then be reduced to $S_a\cdots S_aS_{ab}S_b\cdots S_b\to S_a\cdots S_{aab}S_b\cdots S_b\to S_a\cdots S_{ab}\cdots S_b$, and so on until $S_{ab}$, which is then reduced
to $\bullet$.

    A \emph{finite state machine} is a tuple $A=(\Sigma,Q,q_0,F,\delta)$ where
    \begin{enumerate}
        \item $\Sigma$ is an alphabet,
        \item $Q$ is a set of states,
        \item $q_0\in Q$ is the initial state,
        \item $F\subseteq Q$ is the set of final states,
        \item $\delta\colon Q\times\Sigma\to Q$ is the \emph{transition function}.
    \end{enumerate}
    We define $\delta^*\colon Q\times\Sigma^*\longto Q$ as follows:
    \[ \delta^*(q,\epsilon)=q,\qquad \delta^*(q,\sigma\xi)=\delta^*(\delta(q,\sigma),\xi) \]
    And then we define the \emph{language} of $A$ to be
    \[ \c L(A) = \set{\xi\in\Sigma^*}[\delta^*(q_0,\xi)\in F] \]

    A language $L$ is \emph{regular} if there exists a finite state machine $A$ such that $\c L(A)=L$.
    The class of all regular languages is denoted \REG.

    Every regular language is $\beta$-reducible.

    Let $A=(\Sigma,Q,q_0,F,\delta)$ be a finite state machine.
    We define a \BR-schema $B=(\Sigma,\Gamma,q_0,\blacksquare,\bullet,s,\hb)$ as follows: $\Gamma=\Sigma\cup Q$, $s\colon\sigma\mapsto\sigma$, and for $q\in Q$ and $\sigma\in\Sigma$ we define $\hb(q,\sigma)=\delta(q,\sigma)$.
    And for $q\in F$ we define $\hb(q,\blacksquare)=\bullet$.

    Simple induction will show that $q\xi\to^*\delta^*(q,\xi)$ (where $\xi\to^*\xi'$ means that $\beta^n(\xi)=\xi'$ for some $n$).
    Indeed $\delta^*(q,\epsilon)=q$ and $q\to^*q$, and $q\sigma\xi\to\hb(q,\sigma)\xi=\delta(q,\sigma)\xi$ and by induction $\to^*\delta^*(\delta(q,\sigma),\xi)=\delta^*(q,\sigma\xi)$ as required.
    From this it follows easily that (because $\square=q_0$) $\xi\to^*\delta^*(\xi)$ and thus $\beta^*(\xi)=\bullet$ if and only if $\delta^*(\xi)\in F$, if and only if $\xi\in\c L(A)$.
    Thus $\c L(A)=\c L(B)$ as required.
    \qed

So we have shown that the class of regular languages, \REG, is a proper subclass of \BR{} (since $\set{a^nb^n}[n\geq1]$ is in \BR{} yet not regular).
The question then becomes, what is the relationship between \BR{} and the class of context-free languages \CFL{}, or the class of deterministic context-free languages \DCFL?

\ari{I hope to prove or disprove this (hopefully finding some analogy for the pumping lemma for \BR)}
We conjecture that $\DCFL=\BR$.
For example, we are unsure but believe that the language of even palindromes, $\set{ww^r}[w\in\set{a,b}^+]$, is not \BR.

\ari{The following section is just me writing down some ideas, far from finished}

\subsection{Multi-$\beta$-Reducible Languages}

For $n\geq1$, a $\BR(n)$-schema is a tuple $(\Sigma,\Gamma,\square,\blacksquare,\bullet,s,\set{\hb_i}_{i=0}^{n-1})$ where $\Sigma,\Gamma,\square,\blacksquare,\bullet,s$ are as before and now
\[ \hb_i\colon\Gamma\times(\Gamma\cup\set{\epsilon,\square})\partial(\Gamma\cup\set{\epsilon,\bullet})\times\set{0,\dots,n-1} \]
For each $i\in\set{0,\dots,n-1}$ we define $\beta_i$ as before, but instead of returning the string it also returns the index returned by the reduction.
$\beta^*$ will then start using $\beta_0$, and then when the current $\beta_i$ returns $j$, it switches to $\beta_j$.

We define \MBR{} to be $\MBR=\bigcup_{n\geq1}\BR(n)$.

This is (we believe) stronger than \BR: $\set{ww^r}[w\in\set{a,b}^+]$ is \MBR.
We construct a $\BR(5)$-schema as follows:
\[ B = \parens{\set{a,b},\set{A,B,A',B'},\square,\blacksquare,\bullet,[a\to A,b\to B],\set{\hb_0,\hb_1,\hb_2,\hb_3,\hb_4}} \]
Where we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& \square,A\to A',1\\
        & \square,B\to B',3\\
    \hb_1\colon\quad& A,\square\to A',2\\
    \hb_2\colon\quad& A',A'\to\bullet,2\\
        &A',\epsilon\to\epsilon,2\\
        &A',\square\to A',0\\
    \hb_3\colon\quad& B,\square\to B',2\\
    \hb_4\colon\quad& B',B'\to\bullet,2\\
        &B',\epsilon\to\epsilon,2\\
        &B',\square\to B',0\\
\end{align*}
Intuitively, $\hb_0$ detects the first character and ``tags'' it, then $\hb_1$ or $\hb_3$ ``tag'' the end character if it is the same as the beginning character.
$\hb_2$ and $\hb_4$ ``clean up'' the tags: if the two tagged characters are next to each other, then we can try and accept the word.
Otherwise delete them both.

For the language of palindromes $\set{w\in\set{a,b}^*}[w=w^r]$ we can define a $\BR(3)$-schema where $\bullet=\epsilon$ and
\begin{align*}
    \hb_0\colon\quad& \square,A\to\epsilon,1\\
        & \square,B\to\epsilon,2\\
    \hb_1\colon\quad& A,\blacksquare\to\epsilon,0\\
    \hb_2\colon\quad& B,\blacksquare\to\epsilon,0\\
\end{align*}

But \MBR{} is not contained in \CFL{}: $\set{a^nb^nc^n}[n\geq1]\in\MBR$.
Indeed, we can construct an \MBR-schema as follows:
\[ B = \parens{\set{a,b,c},\set{A,B,C,X},[a\to A,b\to B,c\to C],\set{\hb_0,\hb_1,\hb_2,\hb_3}} \]
And we have the reduction rules:
\begin{align*}
    \hb_0\colon\quad& B,A\to X,0\\
        &C,B\to X,0\\
        &C,\square\to C,1\\
    \hb_1\colon\quad& A,\epsilon\to\epsilon,2\\
    \hb_2\colon\quad& B,\epsilon\to\epsilon,3\\
    \hb_3\colon\quad& C,\square\to\bullet,3\\
        &C,\epsilon\to\epsilon,1
\end{align*}
Intuitively, $\hb_0$ checks that $A$ comes before $B$ which comes before $C$ (as otherwise it places in an $X$, which is never reduced).
Then $\hb_1$ matches a single $A$, then calls $\hb_2$ to match a single $B$, which calls $\hb_3$ to match a single $C$.
